---
title: "The Couch Economy"
subtitle: "Wealth, Inactivity, and the Global NCD Burden"
author: "Ayange Ayangeakaa"
date: today
date-format: "MMMM D, YYYY"
format:
  revealjs:
    theme: [default, custom.scss]
    slide-number: true
    transition: fade
    width: 1280
    height: 720
    embed-resources: true
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
library(dplyr)
library(ggplot2)
library(scales)
library(here)
library(glue)

df <- readRDS(here("data", "processed", "couch_economy.rds"))

income_colors <- c(
  "Low income"          = "#E63946",
  "Lower middle income" = "#F4A261",
  "Upper middle income" = "#2A9D8F",
  "High income"         = "#264653"
)
```

## The Paradox

::: {.columns}

::: {.column width="55%"}

- The world's **wealthiest countries** are often its **most physically inactive**
- High incomes bring sedentary jobs, car-dependent cities, and screen-based leisure
- The result: chronic diseases that cost lives and health systems trillions
- **The Couch Economy** makes this paradox visible and explorable

:::

::: {.column width="45%"}

```{r paradox-scatter, fig.width=5.5, fig.height=4.5}
d <- df |>
  filter(sex == "Both", year == 2022,
         !is.na(inactivity_pct), !is.na(gdp_per_capita), !is.na(income_level))

r <- cor(log10(d$gdp_per_capita), d$inactivity_pct, use = "complete.obs")

ggplot(d, aes(x = gdp_per_capita, y = inactivity_pct, color = income_level)) +
  geom_point(alpha = 0.7, size = 2.5) +
  geom_smooth(aes(group = 1), method = "lm", formula = y ~ x,
              color = "#495057", linewidth = 0.7, linetype = "dashed",
              se = FALSE) +
  scale_x_log10(labels = label_dollar(scale_cut = cut_short_scale())) +
  scale_color_manual(values = income_colors, name = NULL) +
  annotate("text", x = 500, y = 58, label = glue("r = {round(r, 2)}"),
           size = 4, fontface = "italic", color = "#495057", hjust = 0) +
  labs(x = "GDP per capita (log scale)", y = "Physical inactivity (%)") +
  theme_minimal(base_size = 13) +
  theme(legend.position = "bottom",
        panel.grid.minor = element_blank(),
        plot.background = element_rect(fill = "white", colour = NA))
```

:::

:::

## Data Pipeline

```
  [WHO GHO API]     [World Bank API]     [Our World in Data]
  3 indicators       GDP + Gini            Obesity CSV
       |                  |                     |
       +------------------+---------------------+
                          |
                    R/fetch_data.R
                (paginated API calls via httr2)
                          |
                          v  data/raw/*.rds
                          |
                    R/process_data.R
              (join on ISO-3 x year x sex)
                          |
                          v  data/processed/couch_economy.rds
                          |
              +-----------+-----------+
              |                       |
           app.R              R/generate_brief.R
      (Shiny dashboard)       (Gemini API -> brief)
              |
              v
        shinyapps.io
     (refreshed weekly via GitHub Actions)
```

::: {.notes}
Five data sources joined on ISO-3 country code by year and sex. Pipeline handles WHO OData pagination, World Bank Gini sparsity, and OWID CSV format differences.
:::

## Three Capabilities

::: {.columns}

::: {.column width="33%"}

### API Integration

- **WHO GHO** OData API (3 indicators, paginated)
- **World Bank** via `wbstats`
- **OWID** CSV download
- All fetched programmatically at runtime

:::

::: {.column width="33%"}

### GenAI Model

- **Google Gemini** (2.5 Flash Lite) generates country health briefs
- Two audience tones: *policy* and *general public*
- Cached by country + data hash + tone
- Functional, not decorative

:::

::: {.column width="33%"}

### Automation

- **GitHub Actions** cron job every Monday at 06:00 UTC
- Fetches fresh data from all APIs
- Reprocesses, invalidates stale briefs, commits, redeploys
- Zero manual intervention

:::

:::

## Data Wrangling Challenges

| Challenge | Solution |
|-----------|----------|
| WHO, World Bank, and OWID use different country code schemes | `countrycode` package standardises to ISO-3 |
| WHO diabetes API returns two age groups (18+ and 30+), doubling rows on join | Filter to `AGEGROUP_YEARS18-PLUS` before join |
| World Bank Gini index is ~15% missing | Forward/backward fill within country (`tidyr::fill`) |
| WHO regional codes (AFR, AMR, ...) mixed in with country data | `countrycode` returns NA for non-countries; filter them out |
| Venezuela and Ethiopia lack income classifications | Excluded from scatter (uses income colour), kept in trends and briefs |

## Live Demo

::: {.r-fit-text}
[ayange.shinyapps.io/couch-economy](https://ayange.shinyapps.io/couch-economy/)
:::

::: {.notes}
Demo flow:
1. Show default view — scatter plot with trend line, value boxes with correlation
2. Change year, income filter, sex — show reactivity
3. Click a country bubble — show trend chart updating
4. Switch NCD outcome — show diabetes, then mortality
5. Generate a country brief — show policy vs general tone
6. Download a plot and a brief
:::

## Design Decisions

| Decision | Why |
|----------|-----|
| Dual y-axis on trend chart (native plotly) | Inactivity and NCD outcomes have different scales; rescaling to one axis hid true values |
| GDP-inactivity correlation in value box | Communicates the paradox thesis in a single number |
| Cached Gemini briefs (country + hash + tone) | Avoids redundant API calls; instant on repeat visits |
| `bslib` + Bootstrap 5 | Responsive layout, professional polish without custom frontend |
| OLS trend line on scatter | Makes the positive GDP-inactivity relationship immediately visible |

## Limitations and Future Work

**Limitations:**

- WHO inactivity data is **modelled estimates**, not direct measurements — confidence intervals are wide for some countries
- Gini forward-fill assumes inequality changes slowly — may miss abrupt shifts
- Correlation is not causation — the scatter shows association, not a causal pathway

**Future work:**

- Add a **regional comparison** panel (e.g., Sub-Saharan Africa vs. Europe)
- Integrate **WHO STEPS survey** microdata for countries with primary data
- Add **time animation** to the scatter plot to show how the paradox has evolved
- Let users **compare two countries** side by side in the trend chart

## Thank You

::: {.r-fit-text}
[ayange.shinyapps.io/couch-economy](https://ayange.shinyapps.io/couch-economy/)
:::

::: {.columns}

::: {.column width="50%"}

**Source code:**
[github.com/aayange/couch-economy](https://github.com/aayange/couch-economy)

:::

::: {.column width="50%"}

**Course:**
GLHLTH 562 — Spring 2026
Duke University

:::

:::
