---
title: "The Couch Economy"
subtitle: "Wealth, Inactivity, and the Global NCD Burden"
author: "Ayange Ayangeakaa"
date: today
date-format: "MMMM D, YYYY"
format:
  revealjs:
    theme: [default, custom.scss]
    slide-number: true
    transition: fade
    width: 1280
    height: 720
    embed-resources: true
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
library(dplyr)
library(ggplot2)
library(scales)
library(tidyr)
library(here)
library(glue)

df <- readRDS(here("data", "processed", "couch_economy.rds"))

income_colors <- c(
  "Low income"          = "#E63946",
  "Lower middle income" = "#F4A261",
  "Upper middle income" = "#2A9D8F",
  "High income"         = "#264653"
)
```

## The Paradox {.smaller}

::: {.columns}

::: {.column width="50%"}

<br>

> As countries get richer, their populations stop moving.

<br>

- High incomes bring sedentary jobs, car-dependent cities, and screen-based leisure
- The result: rising obesity, diabetes, and premature NCD mortality
- **The Couch Economy** is an interactive dashboard that makes this paradox visible and explorable across 195 countries

:::

::: {.column width="50%"}

```{r paradox-scatter, fig.width=6, fig.height=4.8}
d <- df |>
  filter(sex == "Both", year == 2022,
         !is.na(inactivity_pct), !is.na(gdp_per_capita), !is.na(income_level))

r <- cor(log10(d$gdp_per_capita), d$inactivity_pct, use = "complete.obs")

ggplot(d, aes(x = gdp_per_capita, y = inactivity_pct, color = income_level)) +
  geom_point(alpha = 0.7, size = 2.5) +
  geom_smooth(aes(group = 1), method = "lm", formula = y ~ x,
              color = "#495057", linewidth = 0.8, linetype = "dashed",
              se = FALSE) +
  scale_x_log10(labels = label_dollar(scale_cut = cut_short_scale())) +
  scale_color_manual(values = income_colors, name = NULL) +
  annotate("text", x = 400, y = 60, label = glue("r = {sprintf('%+.2f', r)}"),
           size = 5, fontface = "bold.italic", color = "#495057", hjust = 0) +
  labs(x = "GDP per capita (log scale)", y = "Physical inactivity (%)",
       caption = "Source: WHO GHO + World Bank, 2022") +
  theme_minimal(base_size = 13) +
  theme(legend.position = "bottom",
        panel.grid.minor = element_blank(),
        plot.background = element_rect(fill = "white", colour = NA))
```

:::

:::

::: {.notes}
So here's the big idea. Each dot on this chart is a country in 2022. The x-axis is how rich the country is, the y-axis is how inactive. And the trend line goes up — meaning richer countries actually have more inactive people. You can see countries like Kuwait and the UAE way up at the top right. The r value tells us how strong that relationship is. I built a dashboard that lets you dig into this pattern yourself.
:::

## The Income Gap {.smaller}

::: {.columns}

::: {.column width="50%"}

```{r income-gap, fig.width=6, fig.height=4.8}
gap_data <- df |>
  filter(sex == "Both", !is.na(income_level), !is.na(inactivity_pct)) |>
  group_by(income_level, year) |>
  summarise(avg_inactivity = mean(inactivity_pct, na.rm = TRUE), .groups = "drop")

ggplot(gap_data, aes(x = year, y = avg_inactivity, color = income_level)) +
  geom_line(linewidth = 1.3) +
  geom_point(size = 2) +
  scale_color_manual(values = income_colors, name = NULL) +
  labs(x = NULL, y = "Avg physical inactivity (%)",
       caption = "Source: WHO GHO, 2000-2022") +
  theme_minimal(base_size = 13) +
  theme(legend.position = "bottom",
        panel.grid.minor = element_blank(),
        plot.background = element_rect(fill = "white", colour = NA))
```

:::

::: {.column width="50%"}

<br>

```{r gap-stats}
latest <- gap_data |> filter(year == max(year))
hi <- latest |> filter(income_level == "High income") |> pull(avg_inactivity)
lo <- latest |> filter(income_level == "Low income") |> pull(avg_inactivity)
gap <- round(hi - lo, 1)
```

- High-income countries average **`r round(hi, 1)`%** inactivity
- Low-income countries average **`r round(lo, 1)`%**
- The gap: **`r gap` percentage points**
- And it's been **widening** over two decades
- This isn't random — it's structural

:::

:::

::: {.notes}
Now let me show you how big this gap really is. Each line is an income group. High-income countries are at the top, low-income at the bottom — and they've been pulling apart for twenty years. High-income countries average around 37% inactivity while low-income countries are around 15%. That's a gap of over 20 percentage points, and it's getting worse, not better. This pattern tracks with urbanisation and the shift to desk jobs and car-dependent lifestyles.
:::

## Data Pipeline {.smaller}

::: {.columns}

::: {.column width="55%"}

| Stage | Script | What happens |
|-------|--------|-------------|
| **Ingest** | `R/fetch_data.R` | Fetch 5 indicators from 3 APIs |
| **Process** | `R/process_data.R` | Clean, standardise, join on ISO-3 x year x sex |
| **Serve** | `app.R` | Interactive Shiny dashboard |
| **Generate** | `R/generate_brief.R` | On-demand country briefs via Gemini |
| **Automate** | `.github/workflows/refresh.yml` | Weekly cron: fetch, process, redeploy |

:::

::: {.column width="45%"}

**Data sources:**

| Source | Indicators |
|--------|-----------|
| WHO GHO | Inactivity, diabetes, NCD mortality |
| World Bank | GDP per capita, Gini, income group |
| OWID | Obesity prevalence |

<br>

**Final dataset:**
195 countries, 23 years, 3 sex categories = **13,455 rows**

:::

:::

::: {.notes}
Here's how the data flows. I pull from three different APIs — WHO for the health data, World Bank for the economic data, and Our World in Data for obesity. Nothing is downloaded by hand. The fetch script grabs everything, the process script cleans it and joins it all together using country codes, and then the Shiny app reads that one processed file. The whole pipeline takes about 30 seconds to run. The final dataset is about 13,000 rows covering 195 countries over 23 years.
:::

## Three Capabilities

::: {.columns}

::: {.column width="33%"}

### API Integration

- **WHO GHO** OData API with pagination handler
- **World Bank** via `wbstats` R package
- **OWID** programmatic CSV
- 5 indicators from 3 sources
- No manual downloads

:::

::: {.column width="33%"}

### GenAI Model

- **Google Gemini** 2.5 Flash Lite
- Generates 200-word country health briefs
- Two tones: *policy* vs *general public*
- Cached by country + data hash + tone
- Functional work, not decoration

:::

::: {.column width="33%"}

### Automation

- **GitHub Actions** cron every Monday 06:00 UTC
- Full pipeline: fetch, process, commit, redeploy
- Stale brief cache invalidated automatically
- Zero manual intervention
- Self-healing: rebases on any intervening commits

:::

:::

::: {.notes}
So I'm targeting all three optional capabilities. For APIs — I built a pagination handler for the WHO API since it can return data in chunks. World Bank and OWID are simpler but still fetched programmatically. For the GenAI piece — when you pick a country, the app sends its actual numbers to Google Gemini and gets back a written summary. You can switch between a policy tone and a general public tone. It caches the results so you're not hitting the API twice for the same thing. For automation — there's a GitHub Actions job that runs every Monday. It fetches new data, processes it, and redeploys the app automatically. I don't have to touch anything.
:::

## Data Wrangling Challenges {.smaller}

| Challenge | Solution |
|-----------|----------|
| Three APIs use different country code schemes | `countrycode` package standardises everything to ISO-3 |
| WHO diabetes API returns **two age groups** (18+ and 30+), silently doubling rows on join | Filter to `AGEGROUP_YEARS18-PLUS` before joining |
| World Bank Gini index has ~15% missing values | Forward/backward fill within country (`tidyr::fill`) |
| WHO regional codes (`AFR`, `AMR`, `EMR`, `EUR`, `WPR`) mixed in with country data | `countrycode` returns `NA` for non-countries; filtered out |
| Venezuela and Ethiopia lack World Bank income classifications | Excluded from scatter plot; kept in trends and briefs |
| Trend chart had mismatched scales (inactivity vs NCD outcome) | Rebuilt with native plotly dual y-axis instead of ggplot2 rescaling |

::: {.notes}
This is where a lot of the real work happened. The trickiest one was the diabetes data — the WHO API quietly returns two age groups for every row, 18-plus and 30-plus. I didn't notice at first, and it doubled my entire dataset when I joined. I caught it because the trend chart was showing a fan of lines instead of one clean line. The fix was just one filter, but finding the problem took a while. The Gini data is also tricky because it's really sparse — about 15% missing even after filling gaps. And the dual y-axis issue was that my original chart rescaled everything to one axis, which made the tooltip values wrong.
:::

## Live Demo {.center}

<br><br>

::: {.r-fit-text}
[ayange.shinyapps.io/couch-economy](https://ayange.shinyapps.io/couch-economy/)
:::

<br>

::: {style="text-align: center; color: #6c757d; font-size: 0.9em;"}
Scatter plot | Trend chart | Country briefs | Downloads
:::

::: {.notes}
OK let me show you the app live. Up top you can see the value boxes — the correlation number updates as you change filters. Let me hover over some countries in the scatter plot... each bubble is sized by the NCD outcome. If I click on Kuwait here, the trend chart on the right updates to show Kuwait's data over time. Let me switch the NCD outcome to diabetes — both charts update. Now if I change sex to Female, you can see women are more inactive in most countries. Let me generate a brief for Nigeria — this sends the actual data to Gemini and gets back a written summary. If I switch to general public tone, same data but simpler language. And I can download any of these as a file.
:::

## Design Decisions {.smaller}

| Decision | Rationale |
|----------|-----------|
| Dual y-axis trend chart (native plotly) | Inactivity and NCD outcomes have different scales; rescaling to one axis produced misleading tooltips |
| GDP-inactivity correlation in value box | Communicates the project's thesis in a single number that updates with filters |
| Cached Gemini briefs (country + hash + tone) | Avoids redundant API calls; instant on repeat visits; auto-invalidates when data changes |
| `bslib` + Bootstrap 5 layout | Responsive on mobile and desktop without custom CSS framework |
| OLS trend line on scatter | Makes the positive GDP-inactivity relationship immediately visible to non-technical users |
| Year as dropdown (not slider) | Consistent with other filter inputs; faster to select a specific year |

::: {.notes}
A few of these are worth explaining. The dual y-axis — I originally tried to put both lines on one axis by rescaling, but when you hovered over a point, the numbers were wrong. So I switched to native plotly which lets you have two independent axes. The correlation in the value box — I wanted one number that immediately tells you "yes, richer really does mean more inactive," and it updates when you change filters. The caching — first time you generate a brief it takes a few seconds, but after that it's instant because it's saved locally. And I picked Gemini because it's free and the API is straightforward.
:::

## Limitations and Future Work {.smaller}

::: {.columns}

::: {.column width="50%"}

### Limitations

- WHO inactivity estimates are **modelled**, not directly measured — confidence intervals are wide for some countries
- Gini forward-fill assumes inequality changes slowly — may miss abrupt economic shifts
- **Correlation is not causation** — the scatter shows association, not a causal mechanism
- Gemini briefs can occasionally overinterpret sparse data for small countries

:::

::: {.column width="50%"}

### Future Work

- **Regional comparison panel** — compare Sub-Saharan Africa vs Europe side by side
- **Time animation** on the scatter plot to show how the paradox has evolved since 2000
- **Country comparison mode** — select two countries and see trends overlaid
- **WHO STEPS microdata** integration for countries with primary survey data
- **Confidence interval bands** on the trend chart using WHO's `Low` and `High` columns

:::

:::

::: {.notes}
I want to be upfront about what this can and can't tell you. The WHO numbers are modelled estimates, not direct measurements, so for some smaller countries the uncertainty is pretty wide. And the big one — correlation is not causation. Just because richer countries are more inactive doesn't mean money causes inactivity. There are a lot of other things going on — urbanisation, job types, infrastructure. If I had more time, the thing I'd most want to add is a time animation on the scatter plot so you can watch the dots move over 20 years. That would really make the pattern click.
:::

## Thank You {.center}

<br>

::: {.r-fit-text}
[ayange.shinyapps.io/couch-economy](https://ayange.shinyapps.io/couch-economy/)
:::

<br>

::: {.columns}

::: {.column width="50%" style="text-align: center;"}

**Source code**

[github.com/aayange/couch-economy](https://github.com/aayange/couch-economy)

:::

::: {.column width="50%" style="text-align: center;"}

**Course**

GLHLTH 562 — Spring 2026
Duke University

:::

:::

::: {.notes}
Thanks everyone. The app is live at that link if you want to try it on your own. The code is all on GitHub. Happy to take questions.
:::
